---
title: "4_Alpha_Div_and_barplots"
output: html_document
date: "2025-09-02"
---
```{r libraries and objects}
library("phyloseq"); packageVersion("phyloseq")
library(readr)
library(data.table)
library(ggpubr)
library(ggrepel)
library("ggplot2"); packageVersion("ggplot2")
ps_raw<- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ps_merged_raw.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
ps_0 <- prune_taxa(taxa_sums(ps_raw) > 0, ps_raw)
ps_0
ps_0<- subset_samples(ps_0, Group_cutoff_4 %in% c("R", "NR"))
ps_0<- subset_samples(ps_0, Timepoint %in% c("T0", "T2"))
#target_depth = 3246, identified by  View(table(sample_sums(ps_0)))
ps.rarefied <- rarefy_even_depth(ps_0, sample.size=3246, rngseed=1, replace=FALSE)
```

#Taxonomy

Alpha-diversity is calculated on the raw data, here data_otu or data_phylo if you are using phyloseq.
It is important to not use filtered data because many richness estimates are modeled on singletons and doubletons in the occurrence table. So, you need to leave them in the dataset if you want a meaningful estimate.
Moreover, we usually not using normalized data because we want to assess the diversity on the raw data and we are not comparing Volunteer to each other but only assessing diversity within each sample.

```{r Counts}
reads_per_sample <- as.matrix(rowSums(t(as.data.frame(otu_table(ps_0)))))
reads_per_sample 
```

```{r Alpha Diversity}

library(ggsignif)
#Shannon
Shannon<-phyloseq::estimate_richness(ps_0, measures = "Shannon")
metadata<- data.frame(sample_data(ps_0)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon, aes(x = GroupTime, y = Shannon)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha =1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")

#Chao1
Chao1<-phyloseq::estimate_richness(ps_0, measures = "Chao1")
metadata<- data.frame(sample_data(ps_0)) 
Chao1$Timepoint<-metadata$Timepoint
Chao1$Group_cutoff_4<-metadata$Group_cutoff_4
Chao1$Volunteer<-metadata$Volunteer
Chao1$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Chao1$GroupTime <-metadata$GroupTime 
pChao1<- ggplot(Chao1, aes(x = GroupTime, y = Chao1)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha =1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Chao1 Index")

#Fisher
Fisher<-phyloseq::estimate_richness(ps_0, measures = "Fisher")
metadata<- data.frame(sample_data(ps_0))
Fisher$Timepoint<-metadata$Timepoint
Fisher$Group_cutoff_4<-metadata$Group_cutoff_4
Fisher$Volunteer<-metadata$Volunteer
Fisher$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Fisher$GroupTime <-metadata$GroupTime 
pFisher<- ggplot(Fisher, aes(x = GroupTime, y = Fisher)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha =1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Fisher Index")


ggarrange(pShannon, pFisher, pChao1, common.legend = TRUE, legend="bottom",nrow = 1)
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_all.pdf", height = 5, width = 7.5)
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_all.svg", height = 5, width = 7.5)
#Error in loadNamespace(x) : there is no package called ‘svglite’

```


```{r comaprisons for significance}
library(lme4)
library(lmerTest)   # for Type III ANOVA with Satterthwaite
library(emmeans)
library(openxlsx)

# Create a new workbook
wb <- createWorkbook()

# List of datasets and their names
datasets <- list(Shannon=Shannon, Fisher=Fisher, Chao1=Chao1)
response_vars <- c("Shannon", "Fisher", "Chao1")

for(i in seq_along(datasets)) {
    data <- datasets[[i]]
    response <- response_vars[i]
    
    # Fit model
    model <- lmer(as.formula(paste(response, "~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer)")), 
                  data = data)
    
    # ANOVA table
    anova_tab <- as.data.frame(anova(model))
    sheet_name_anova <- paste0(response, "_ANOVA")
    addWorksheet(wb, sheet_name_anova)
    writeData(wb, sheet_name_anova, anova_tab, rowNames = TRUE)
    
    # emmeans: Group_cutoff_4 | Timepoint
    emm1 <- emmeans(model, ~ Group_cutoff_4 | Timepoint)
    emm1_tab <- as.data.frame(pairs(emm1))
    sheet_name_emm1 <- paste0(response, "_G_by_T")
    addWorksheet(wb, sheet_name_emm1)
    writeData(wb, sheet_name_emm1, emm1_tab)
    
    # emmeans: Timepoint | Group_cutoff_4
    emm2 <- emmeans(model, ~ Timepoint | Group_cutoff_4)
    emm2_tab <- as.data.frame(pairs(emm2))
    sheet_name_emm2 <- paste0(response, "_T_by_G")
    addWorksheet(wb, sheet_name_emm2)
    writeData(wb, sheet_name_emm2, emm2_tab)
    
    # emmeans: Timepoint (main effect)
    emm3 <- emmeans(model, ~ Timepoint)
    emm3_tab <- as.data.frame(pairs(emm3))
    sheet_name_emm3 <- paste0(response, "_T_main")
    addWorksheet(wb, sheet_name_emm3)
    writeData(wb, sheet_name_emm3, emm3_tab)
}

# Save workbook
saveWorkbook(wb, "/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Tables/ANOVA_EMM_results.xlsx", overwrite = TRUE)
```

###Barplots
```{r barplots}
ps_merged_filtered <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ps_merged_filtered.Rds")
# Load required libraries
library(phyloseq)
library(ggplot2)
library(RColorBrewer)

ps_merged_filtered<- subset_samples(ps_merged_filtered, Group_cutoff_4 %in% c("R", "NR"))
ps_merged_filtered<- subset_samples(ps_merged_filtered, Timepoint %in% c("T0", "T2"))

# Agglomerate at Phylum level
ps_phylum <- tax_glom(ps_merged_filtered, taxrank = "Phylum")

# Transform to relative abundance
# Convert to percentages (0–100 scale)
ps_rel <- transform_sample_counts(ps_phylum, function(x) 100 * x / sum(x))


# Melt to long format
df_long <- psmelt(ps_rel)

# Optional: Tidy Phylum names (e.g., replace NAs or simplify)
df_long$Phylum <- as.character(df_long$Phylum)
df_long$Phylum[is.na(df_long$Phylum)] <- "Unclassified"

# Set order of phyla by abundance (optional)
top_phyla <- names(sort(tapply(df_long$Abundance, df_long$Phylum, sum), decreasing=TRUE))
df_long$Phylum <- factor(df_long$Phylum, levels = top_phyla)

df_summary <- df_long %>%
    group_by(Group_cutoff_4, Phylum) %>%
    summarise(Abundance = mean(Abundance), .groups = "drop")

df_summary <- df_summary %>%
    mutate(Phylum = case_when(
        Phylum %in% c("NA unknown", "NA Bacteria phylum incertae sedis") ~ "Unclassified",
        TRUE ~ as.character(Phylum)
    ))

# Optional: reorder phyla
top_phyla <- names(sort(tapply(df_summary$Abundance, df_summary$Phylum, sum), decreasing = TRUE))
df_summary$Phylum <- factor(df_summary$Phylum, levels = top_phyla)

# Plot
color_palette <- c(
    "Bacteroidetes" = "#0072B2",    # dark blue
    "Firmicutes" = "#56B4E9",     # light blue 
    "Proteobacteria" = "#F0E442",    # yellow 
    "Actinobacteria" = "#CC79A7",    # pink
    "Elusimicrobia" = "#D55E00",    # dark orange 
    "Tenericutes" = "#E69F00",        # orange 
    "Ignavibacteriae" = "#009E73",    # green
    "Unclassified" = "#999999"       # grey
)


ggplot(df_summary, aes(x = Group_cutoff_4, y = Abundance, fill = Phylum)) +
    geom_bar(stat = "identity", position = "stack", alpha = 0.6) +
    scale_fill_manual(values = color_palette) +
    theme_classic(base_size = 14) +
    labs(x = "Group", y = "Relative Abundance (%)", fill = "Phylum") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ylim(0, 100)

ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Barplots_tax.pdf", height = 5, width = 4)
```

###### Significant diff barplots
```{r lmms}
#change to clr

library(phyloseq)
library(compositions)

ps_phylum <- tax_glom(ps_merged_filtered, taxrank = "Phylum")

otu <- as.data.frame(t(otu_table(ps_phylum)))
otu_clr <- clr(otu + 1)
otu_table(ps_phylum) <- otu_table(t(otu_clr), taxa_are_rows = TRUE)

df_lmm <- psmelt(ps_phylum)

df_lmm <- df_lmm %>%
  mutate(
    Phylum = as.character(Phylum),
    Timepoint = factor(Timepoint, levels = c("T0", "T2")),
    Group_cutoff_4 = factor(Group_cutoff_4, levels = c("NR", "R"))
  )

# Run LMM for each phylum
library(lme4)
library(lmerTest)
library(broom.mixed)

lmm_results <- df_lmm %>%
  group_by(Phylum) %>%
  do(
    tidy(
      lmer(
        Abundance ~ Group_cutoff_4 * Timepoint + (1 | Volunteer),
        data = .
      )
    )
  ) %>%
  ungroup()

interaction_results <- lmm_results %>%
  filter(term == "Group_cutoff_4R:TimepointT2") %>%
  mutate(p.adj = p.adjust(p.value, method = "BH"))

#Significant differences indicate that R and NR change differently from T0 to T2

# A tibble: 9 × 10
#  Phylum      effect group term  estimate std.error statistic    df p.value  p.adj
#  <chr>       <chr>  <chr> <chr>    <dbl>     <dbl>     <dbl> <dbl>   <dbl>  <dbl>
#Actinobacteria fixed  NA    Grou…   -1.49      0.787    -1.89   13.1 0.0808  0.170
#Bacteroidetes  fixed  NA    Grou…    2.20      1.21      1.82   11.4 0.0946  0.170
#Elusimicrobia  fixed  NA    Grou…    1.38      0.925     1.50   31.0 0.144   0.186
#Firmicutes   fixed  NA    Grou…   -1.02      0.351    -2.89   31.0 0.00690 0.0563
#Ignavibacteriae fixed NA  Grou…   -0.750     0.388    -1.93   15.4 0.0717  0.170 
#NA Bacteria phy fixed  NA   Grou…    0.203     0.382     0.532  15.5 0.602   0.602 #NA unknown    fixed  NA    Grou…   -0.819     0.287    -2.85   14.4 0.0125  0.0563
#Proteobacteria fixed  NA    Grou…   -0.802     0.624    -1.29   15.0 0.218   0.245
#Tenericutes   fixed  NA    Grou…    1.21      0.766     1.58   15.7 0.134   0.186

#All groups T0 vs T2
time_results <- lmm_results %>%
  filter(term == "TimepointT2") %>%
  mutate(p.adj = p.adjust(p.value, method = "BH"))
#All Tiempoints R vs NR
group_results <- lmm_results %>%
  filter(term == "Group_cutoff_4R") %>%
  mutate(p.adj = p.adjust(p.value, method = "BH"))
#model diagnostics
fit <- lmer(
  Abundance ~ Group_cutoff_4 * Timepoint + (1 | Volunteer),
  data = filter(df_lmm, Phylum == "Firmicutes")
)

plot(fit)
qqnorm(residuals(fit))
qqline(residuals(fit))


#Helper function: LMM + emmeans for all phyla
library(lme4)
library(lmerTest)
library(emmeans)
library(dplyr)
library(purrr)
library(broom)

run_emmeans_lmm <- function(
  data,
  response = "Abundance",
  phylum_col = "Phylum",
  subject_col = "Volunteer",
  group_col = "Group_cutoff_4",
  time_col = "Timepoint",
  p_adjust = "BH"
) {

  # Ensure factors are ordered correctly
  data <- data %>%
    mutate(
      !!group_col := factor(.data[[group_col]], levels = c("NR", "R")),
      !!time_col  := factor(.data[[time_col]],  levels = c("T0", "T2"))
    )

  # Nest by phylum
  nested <- data %>%
    group_by(.data[[phylum_col]]) %>%
    nest()

  # Fit models + emmeans
  results <- nested %>%
    mutate(
      model = map(
        data,
        ~ lmer(
          as.formula(
            paste(response, "~", group_col, "*", time_col, "+ (1 |", subject_col, ")")
          ),
          data = .
        )
      ),

      # Within-group T0 vs T2
      within_group = map(
        model,
        ~ emmeans(., pairwise ~ !!sym(time_col) | !!sym(group_col))$contrasts %>%
          as.data.frame() %>%
          mutate(contrast_type = "Within-group (T0 vs T2)")
      ),

      # Between-group R vs NR at each timepoint
      between_group = map(
        model,
        ~ emmeans(., pairwise ~ !!sym(group_col) | !!sym(time_col))$contrasts %>%
          as.data.frame() %>%
          mutate(contrast_type = "Between-group (R vs NR)")
      )
    )

  # Combine results
  out <- results %>%
    select(.data[[phylum_col]], within_group, between_group) %>%
    pivot_longer(
      cols = c(within_group, between_group),
      names_to = "comparison",
      values_to = "results"
    ) %>%
    unnest(results) %>%
    group_by(contrast_type) %>%
    mutate(p.adj = p.adjust(p.value, method = p_adjust)) %>%
    ungroup()

  return(out)
}

emmeans_results <- run_emmeans_lmm(df_lmm)

#Firmicutes/Bacteroides ratio difefrences
df_fb <- df_lmm %>%
    filter(Phylum %in% c("Firmicutes", "Bacteroidetes")) %>%
    select(Volunteer, Timepoint, Group_cutoff_4, Phylum, Abundance) %>%
    pivot_wider(names_from = Phylum, values_from = Abundance) %>%
    mutate(FB_ratio = Firmicutes - Bacteroidetes)

fb_model <- lmer(FB_ratio ~ Group_cutoff_4 * Timepoint + (1 | Volunteer), data = df_fb)

anova(fb_model)  # Type III ANOVA

emmeans(fb_model, pairwise ~ Group_cutoff_4 | Timepoint) %>%
    .$contrasts %>%
    as.data.frame()%>%
    mutate(p.adj = p.adjust(p.value, method = "BH"))

emmeans(fb_model, pairwise ~ Timepoint | Group_cutoff_4) %>%
    .$contrasts %>%
    as.data.frame()%>%
    mutate(p.adj = p.adjust(p.value, method = "BH"))
# contrast Group_cutoff_4  estimate        SE       df   t.ratio   p.value     p.adj
#T0 - T2             NR -1.323030 0.8526291 15.87433 -1.551706 0.1404391 0.1404391
#T0 - T2              R  1.942295 1.1730271 15.19796  1.655797 0.1182598 0.1404391



#Firmicutes/Bacteroides ratio difefrences
df_fb <- df_lmm %>%
    filter(Phylum %in% c("Firmicutes", "Bacteroidetes")) %>%
    select(Volunteer, Timepoint, Group_cutoff_4, Phylum, Abundance) %>%
    pivot_wider(names_from = Phylum, values_from = Abundance) %>%
    mutate(FB_ratio = Firmicutes - Bacteroidetes)

fb_model <- lmer(FB_ratio ~ Group_cutoff_4 * Timepoint + (1 | Volunteer), data = df_fb)

anova(fb_model)  # Type III ANOVA

emmeans(fb_model, pairwise ~ Group_cutoff_4 | Timepoint) %>%
    .$contrasts %>%
    as.data.frame()%>%
    mutate(p.adj = p.adjust(p.value, method = "BH"))

emmeans(fb_model, pairwise ~ Timepoint | Group_cutoff_4) %>%
    .$contrasts %>%
    as.data.frame()%>%
    mutate(p.adj = p.adjust(p.value, method = "BH"))
# contrast Group_cutoff_4  estimate        SE       df   t.ratio   p.value     p.adj
#T0 - T2             NR -1.323030 0.8526291 15.87433 -1.551706 0.1404391 0.1404391
#T0 - T2              R  1.942295 1.1730271 15.19796  1.655797 0.1182598 0

```


#GMMs

```{r libraries and objects GMMs}
library("phyloseq"); packageVersion("phyloseq")
library(readr)
library(data.table)
library(ggpubr)
library(ggrepel)
library("ggplot2"); packageVersion("ggplot2")
GMMs_corrected<- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/GMMs_corrected.Rds")
# Because ARSyN applies log transformation we have negative values, to perform shannon we need no negative entries
otu<-abundances(GMMs_corrected)+135911
otu_table(GMMs_corrected) <- otu_table(otu, taxa_are_rows = TRUE)

GMM_raw<- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/GMMs_merged.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
GMM_0 <- prune_taxa(taxa_sums(GMM_raw) > 0, GMM_raw)
GMM_0
GMM_0<- subset_samples(GMM_0, Group_cutoff_4 %in% c("R", "NR"))
GMM_0<- subset_samples(GMM_0, Timepoint %in% c("T0", "T2"))

```

```{r GMM Aplha Diversity}

library(ggsignif)
#Shannon
Shannon<-phyloseq::estimate_richness(GMM_0, measures = "Shannon")
metadata<- data.frame(sample_data(GMM_0)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon, aes(x = GroupTime, y = Shannon)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha =1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")
pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_GMM_0.pdf", height = 5, width = 4)


#Shannon corrected
Shannon_corrected<-phyloseq::estimate_richness(GMMs_corrected, measures = "Shannon")
metadata<- data.frame(sample_data(GMMs_corrected)) 
Shannon_corrected$Timepoint<-metadata$Timepoint
Shannon_corrected$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon_corrected$Volunteer<-metadata$Volunteer
Shannon_corrected$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon_corrected$GroupTime <-metadata$GroupTime 

pShannon <- ggplot(Shannon_corrected, aes(x = GroupTime, y = Shannon)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha = 1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index") +
    
    # Add significance stars
    geom_signif(
        comparisons = list(c("NR_T0", "R_T0"), c("NR_T2", "R_T2")),
        annotations = c("***", "***"),
        y_position = c(max(Shannon_corrected$Shannon) + 0.01, max(Shannon_corrected$Shannon) + 0.03),
        tip_length = 0.03,
        textsize = 5
    )

pShannon

ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_GMMs_corrected.pdf", height = 5, width = 4)


# Extract OTU table - samples in columns or rows (transpose if needed)
otu_counts <- as.data.frame(t(otu_table(GMM_0)))
# Calculate total reads per sample
total_reads <- rowSums(otu_counts)

# Create data frame and add metadata
metadata <- data.frame(sample_data(GMM_0))
total_reads_df <- data.frame(Sample = rownames(metadata), TotalReads = total_reads)
total_reads_df$Timepoint <- metadata$Timepoint
total_reads_df$Group_cutoff_4 <- metadata$Group_cutoff_4
total_reads_df$Volunteer <- metadata$Volunteer
total_reads_df$Study <- metadata$Study

# Create combined group variable if needed
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
total_reads_df$GroupTime <- metadata$GroupTime

# Plot total reads with ggplot2
pTotalReads <- ggplot(total_reads_df, aes(x = GroupTime, y = TotalReads)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha = 1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Total Reads")
pTotalReads
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Total_reads_GMMs_0.pdf", height = 5, width = 4)

```


```{r comaprisons for significance}
Shannon<-phyloseq::estimate_richness(GMM_0, measures = "Shannon")
metadata<- data.frame(sample_data(GMM_0)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 

model<-lmer(Shannon ~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer),data=Shannon)
anova(model)
#Type III Analysis of Variance Table with Satterthwaite's method
#                                  Sum Sq   Mean Sq NumDF  DenDF  F value    Pr(>F)    
#Timepoint                      0.0000031 0.0000031     1 13.265   0.0255   0.87546    
#Group_cutoff_4                 0.0164974 0.0164974     1 13.648 134.0642 1.934e-08 ***
#Study                          0.0008418 0.0008418     1 13.648   6.8411   0.02069 *  
#Timepoint:Group_cutoff_4       0.0000276 0.0000276     1 13.265   0.2242   0.64359    
#Timepoint:Study                0.0002211 0.0002211     1 13.265   1.7969   0.20260    
#Group_cutoff_4:Study           0.0000238 0.0000238     1 13.648   0.1938   0.66669    
#Timepoint:Group_cutoff_4:Study 0.0001218 0.0001218     1 13.265   0.9899   0.33758    
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


emm <- emmeans(model, ~ Group_cutoff_4 | Timepoint )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Timepoint = T0:
# contrast estimate      SE   df t.ratio p.value
# NR - R    -0.0735 0.00747 23.8  -9.842 <0.0001

#Timepoint = T2:
# contrast estimate      SE   df t.ratio p.value
# NR - R    -0.0696 0.00743 23.7  -9.372 <0.0001

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 

emm <- emmeans(model, ~ Timepoint | Group_cutoff_4 )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Group_cutoff_4 = NR:
# contrast estimate      SE   df t.ratio p.value
# T0 - T2  -0.00130 0.00529 13.3  -0.246  0.8094

#Group_cutoff_4 = R:
# contrast estimate      SE   df t.ratio p.value
# T0 - T2   0.00263 0.00640 13.1   0.411  0.6881

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Timepoint  )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
# contrast estimate      SE   df t.ratio p.value
# T0 - T2  0.000663 0.00415 13.2   0.160  0.8756

#Results are averaged over the levels of: Group_cutoff_4, Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Group_cutoff_4  )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
# contrast estimate      SE   df t.ratio p.value
# NR - R    -0.0716 0.00618 13.6 -11.576 <0.0001

#Results are averaged over the levels of: Timepoint, Study 
#Degrees-of-freedom method: kenward-roger 

# Perform t-tests for the specific pairs
t1 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "NR_T2"])
t2 <- t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])
t3<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "UK"])
t4<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "UK"])
t5<-t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "R_T0" & Shannon$Study == "UK"])
t6<-t.test(Shannon$Shannon[Shannon$GroupTime == "R_T2" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "R_T2" & Shannon$Study == "UK"])
t7 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T0"])
t8 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
    comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "R_T0 Spain vs R_T0 UK", "R_T2 Spain vs R_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
    p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value, t6$p.value, t7$p.value, t8$p.value)
)

# Adjust p-values using BH method
test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon$signif <- cut(
  test_results_Sannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)
test_results_Sannon

#> test_results_Shannon
#               comparison      p_value       p_adj signif
#1          NR_T0 vs NR_T2 0.5957623119 0.595762312       
#2            R_T0 vs R_T2 0.4381242880 0.584165717       
#3 NR_T0 Spain vs NR_T0 UK 0.0004512338 0.003609870     **
#4 NR_T2 Spain vs NR_T2 UK 0.0058843946 0.015691719      *
#5   R_T0 Spain vs R_T0 UK 0.0473265338 0.094653068       
#6   R_T2 Spain vs R_T2 UK 0.0011916085 0.004766434     **
#7           NR_T0 vs R_T0 0.0916416743 0.146626679       
#8           NR_T2 vs R_T2 0.5514133827 0.595762312  

# Perform t-tests for the specific pairs
t1 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"])
t2 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])
t3<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "UK"])
t4<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "UK"])
t5<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0" & Shannon_corrected$Study == "UK"])
t6<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2" & Shannon_corrected$Study == "UK"])
t7 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"])
t8 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon_corrected <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "R_T0 Spain vs R_T0 UK", "R_T2 Spain vs R_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value, t6$p.value, t7$p.value, t8$p.value)
)

# Adjust p-values using BH method
test_results_Sannon_corrected$p_adj <- p.adjust(test_results_Sannon_corrected$p_value, method = "BH")
test_results_Sannon_corrected$signif <- cut(
  test_results_Sannon_corrected$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)
test_results_Sannon_corrected

#> test_results_Shannon_corrected
#               comparison      p_value        p_adj signif
#1          NR_T0 vs NR_T2 7.601401e-01 8.355632e-01       
#2            R_T0 vs R_T2 7.945825e-01 8.355632e-01       
#3 NR_T0 Spain vs NR_T0 UK 3.787914e-01 6.060662e-01       
#4 NR_T2 Spain vs NR_T2 UK 2.757410e-03 7.353093e-03     **
#5   R_T0 Spain vs R_T0 UK 8.355632e-01 8.355632e-01       
#6   R_T2 Spain vs R_T2 UK 9.648041e-02 1.929608e-01       
#7           NR_T0 vs R_T0 2.147938e-05 8.591752e-05    ***
#8           NR_T2 vs R_T2 1.335429e-05 8.591752e-05    ***
```

###Barplots
```{r barplots}
GMM_raw<- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/GMMs_merged.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
GMM_0 <- prune_taxa(taxa_sums(GMM_raw) > 0, GMM_raw)
GMM_0
GMM_0<- subset_samples(GMM_0, Group_cutoff_4 %in% c("R", "NR"))
GMM_0<- subset_samples(GMM_0, Timepoint %in% c("T0", "T2"))# Load required libraries
tax<-as.data.frame(tax_table(GMM_0))

tax <- tax %>%
  mutate(
    Category = case_when(
      grepl("degradation", V2, ignore.case = TRUE) & grepl("tyrosine|phenylalanine|tryptophan|glutamate|alanine|valine|isoleucine|leucine|methionine|cysteine|aspartate|asparagine|glycine|serine|threonine|proline|histidine|lysine|arginine|glutamine", V2) ~ "Amino Acid Degradation",
      grepl("synthesis", V2, ignore.case = TRUE) & grepl("tyrosine|phenylalanine|tryptophan|glutamate|GABA|serotonin|histamine|dopamine", V2) ~ "Amino Acid or Neuroactive Synthesis",
      grepl("degradation", V2, ignore.case = TRUE) & grepl("lactose|sucrose|fructose|galactose|starch|trehalose|arabinose|xylose|mannose|rhamnose|melibiose|maltose|allose|pectin|fructan|galacturonate", V2) ~ "Sugar Degradation",
      grepl("glycolysis|pentose phosphate|Entner-Doudoroff|Bifidobacterium shunt", V2, ignore.case = TRUE) ~ "Sugar Metabolism",
      grepl("butyrate|propionate|acetate", V2, ignore.case = TRUE) ~ "SCFA Metabolism",
      grepl("menaquinone|SAM|vitamin", V2, ignore.case = TRUE) ~ "Vitamin & Cofactor Metabolism",
      grepl("serotonin|GABA|dopamine|histamine|tryptamine", V2, ignore.case = TRUE) ~ "Neuroactive Compound Metabolism",
      grepl("triacylglycerol|glycerol|fatty acid", V2, ignore.case = TRUE) ~ "Lipid Metabolism",
      grepl("nitric oxide|hydrogen|sulfate|nitrate", V2, ignore.case = TRUE) ~ "Gas Metabolism",
      grepl("methanogenesis", V2, ignore.case = TRUE) ~ "Methanogenesis",
      grepl("quinolinic acid|DOPAC|NAD", V2, ignore.case = TRUE) ~ "Nucleotide/Quinone Metabolism",
      grepl("chaperone", V2, ignore.case = TRUE) ~ "Peptide/Protein Metabolism",
      TRUE ~ "Other"
    ),
    Category_Description = case_when(
      Category == "Amino Acid Degradation" ~ "Breakdown of amino acids like tyrosine, glutamate, lysine, etc.",
      Category == "Amino Acid or Neuroactive Synthesis" ~ "Biosynthesis of amino acids and related neurotransmitters (e.g., GABA, serotonin).",
      Category == "Sugar Degradation" ~ "Catabolism of mono-, di-, and polysaccharides (e.g., lactose, pectin, trehalose).",
      Category == "Sugar Metabolism" ~ "Core sugar metabolic pathways like glycolysis and pentose phosphate pathway.",
      Category == "SCFA Metabolism" ~ "Production or degradation of short-chain fatty acids (butyrate, propionate, acetate).",
      Category == "Vitamin & Cofactor Metabolism" ~ "Biosynthesis of vitamins and cofactors like vitamin K2, SAM.",
      Category == "Neuroactive Compound Metabolism" ~ "Metabolism of compounds affecting the nervous system (e.g., GABA, dopamine).",
      Category == "Lipid Metabolism" ~ "Metabolism of lipids, glycerol, and fatty acids.",
      Category == "Gas Metabolism" ~ "Involves nitrogen, sulfur, and hydrogen compound transformations (e.g., NO, sulfate).",
      Category == "Methanogenesis" ~ "Microbial pathways producing methane from carbon or methyl sources.",
      Category == "Nucleotide/Quinone Metabolism" ~ "Metabolism of nucleotide-like or redox-active compounds (e.g., quinolinic acid).",
      Category == "Peptide/Protein Metabolism" ~ "Chaperone proteins and protein folding/degradation.",
      Category == "Other" ~ "Does not fit clearly into predefined categories.",
      TRUE ~ NA_character_
    )
  )
tax_table(GMM_0) <- tax_table(as.matrix(tax))


library(phyloseq)
library(ggplot2)
library(RColorBrewer)

# Agglomerate at Category level
GMM_Category <- tax_glom(GMM_0, taxrank = "Category")

# Melt to long format
df_long <- psmelt(GMM_0)

# Optional: Tidy Category names (e.g., replace NAs or simplify)
df_long$Category <- as.character(df_long$Category)
df_long$Category[is.na(df_long$Category)] <- "Unclassified"
library(dplyr)
library(ggplot2)

# Step 1: Calculate mean abundance per group and category
df_summary <- df_long %>%
  group_by(Group_cutoff_4, Category) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

# Step 2: Normalize within each group to make each bar total 100%
df_summary <- df_summary %>%
  group_by(Group_cutoff_4) %>%
  mutate(RelativeAbundance = 100 * Abundance / sum(Abundance)) %>%
  ungroup()

# Step 3: Reorder categories by total relative abundance (optional)
top_categories <- names(sort(tapply(df_summary$RelativeAbundance, df_summary$Category, sum), decreasing = TRUE))
df_summary$Category <- factor(df_summary$Category, levels = top_categories)

# Step 4: Plot
ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Category)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Category") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)


ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Barplots_GMMs.pdf", height = 5, width = 5)
```

###### Significant diff barplots GMM
```{r GMMMs barplots}
df_lmm <- df_long %>%
  filter(Timepoint %in% c("T0", "T2"),
         Group_cutoff_4 %in% c("R", "NR")) %>%
  select(Volunteer, Timepoint, Group_cutoff_4, Category, Abundance) %>%
  rename(Phylum = Category)  # so we can reuse the LMM code

emmeans_results <- run_emmeans_lmm(df_lmm)

```


#ARGs ok
```{r libraries and objects ARGs}
library("phyloseq"); packageVersion("phyloseq")
library(readr)
library(data.table)
library(ggpubr)
library(ggrepel)
library("ggplot2"); packageVersion("ggplot2")
ARGs_ok_corrected <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_ok_corrected.Rds")
# Because ARSyN applies log transformation we have negative values, to perform shannon we need no negative entries
otu<-abundances(ARGs_ok_corrected)+2607523
otu_table(ARGs_ok_corrected) <- otu_table(otu, taxa_are_rows = TRUE)

ARGs_ok <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_ok.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
ARGs_ok<- subset_samples(ARGs_ok, Group_cutoff_4 %in% c("R", "NR"))
ARGs_ok<- subset_samples(ARGs_ok, Timepoint %in% c("T0", "T2"))

```

```{r ARRs ok Aplha Diversity}

library(ggsignif)
#Shannon
Shannon<-phyloseq::estimate_richness(ARGs_ok, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ok)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon, aes(x = GroupTime, y = Shannon)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha =1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")
pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_ARGs_ok.pdf", height = 5, width = 4)


#Shannon corrected
Shannon_corrected<-phyloseq::estimate_richness(ARGs_ok_corrected, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ok_corrected)) 
Shannon_corrected$Timepoint<-metadata$Timepoint
Shannon_corrected$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon_corrected$Volunteer<-metadata$Volunteer
Shannon_corrected$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon_corrected$GroupTime <-metadata$GroupTime 

pShannon<- ggplot(Shannon_corrected, aes(x = GroupTime, y = Shannon)) +
    geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
    geom_jitter(aes(color = Study), alpha =1,
                position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
    scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
    scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
    theme_minimal() +
    labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index") +
    geom_signif(
        comparisons = list(c("NR_T0", "R_T0")),
        annotations = c("*"),
        y_position = c(max(Shannon_corrected$Shannon) + 0.000000001),
        tip_length = 0.03,
        textsize = 5
    )

pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_ARGs_ok_corrected.pdf", height = 5, width = 4)


# Extract OTU table - samples in columns or rows (transpose if needed)
otu_counts <- as.data.frame(t(otu_table(ARGs_ok)))
# Calculate total reads per sample
total_reads <- rowSums(otu_counts)

# Create data frame and add metadata
metadata <- data.frame(sample_data(ARGs_ok))
total_reads_df <- data.frame(Sample = rownames(metadata), TotalReads = total_reads)
total_reads_df$Timepoint <- metadata$Timepoint
total_reads_df$Group_cutoff_4 <- metadata$Group_cutoff_4
total_reads_df$Volunteer <- metadata$Volunteer
total_reads_df$Study <- metadata$Study

# Create combined group variable if needed
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
total_reads_df$GroupTime <- metadata$GroupTime

# Plot total reads with ggplot2
pTotalReads <- ggplot(total_reads_df, aes(x = GroupTime, y = TotalReads)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha = 1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Total Reads")
pTotalReads
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Total_reads_ARG_ok.pdf", height = 5, width = 4)

```


```{r comaprisons for significance ARG ok}
Shannon<-phyloseq::estimate_richness(ARGs_ok, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ok)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 

model<-lmer(Shannon ~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer),data=Shannon)
#boundary (singular) fit: see help('isSingular')
anova(model)
#Type III Analysis of Variance Table with Satterthwaite's method
#                               Sum Sq Mean Sq NumDF DenDF F value    Pr(>F)    
#Timepoint                      0.4311  0.4311     1    15  1.1209 0.3064795    
#Group_cutoff_4                 1.9061  1.9061     1    15  4.9563 0.0417459 *  
#Study                          9.4773  9.4773     1    15 24.6429 0.0001698 ***
#Timepoint:Group_cutoff_4       0.9793  0.9793     1    15  2.5464 0.1313924    
#Timepoint:Study                0.5710  0.5710     1    15  1.4847 0.2418569    
#Group_cutoff_4:Study           0.3183  0.3183     1    15  0.8277 0.3773123    
#Timepoint:Group_cutoff_4:Study 0.5242  0.5242     1    15  1.3629 0.2612524    
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

emm <- emmeans(model, ~ Group_cutoff_4 | Timepoint )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Timepoint = T0:
# contrast estimate    SE df t.ratio p.value
# NR - R     -1.170 0.449 15  -2.605  0.0199

#Timepoint = T2:
# contrast estimate    SE df t.ratio p.value
# NR - R     -0.193 0.425 15  -0.454  0.6561

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Timepoint | Group_cutoff_4 )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Group_cutoff_4 = NR:
# contrast estimate    SE   df t.ratio p.value
# T0 - T2    -0.164 0.355 8.36  -0.463  0.6549

#Group_cutoff_4 = R:
# contrast estimate    SE   df t.ratio p.value
# T0 - T2     0.813 0.506 6.69   1.605  0.1545

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Timepoint  )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
# contrast estimate    SE   df t.ratio p.value
# T0 - T2     0.324 0.309 7.18   1.049  0.3284

#Results are averaged over the levels of: Group_cutoff_4, Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Group_cutoff_4  )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
# contrast estimate    SE   df t.ratio p.value
# NR - R     -0.682 0.309 7.18  -2.205  0.0623

#Results are averaged over the levels of: Timepoint, Study 
#Degrees-of-freedom method: kenward-roger

# Perform t-tests for the specific pairs
t1 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "NR_T2"])
t2 <- t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])
t3<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "UK"])
t4<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "UK"])

t7 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T0"])
t8 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t7$p.value, t8$p.value)
)

test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon$signif <- cut(
  test_results_Sannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

test_results_Sannon
#               comparison    p_value     p_adj signif
#1          NR_T0 vs NR_T2 0.55468496 0.57820608       
#2            R_T0 vs R_T2 0.57820608 0.57820608       
#3 NR_T0 Spain vs NR_T0 UK 0.05159218 0.10318435       
#4 NR_T2 Spain vs NR_T2 UK 0.02486760 0.07460281       
#5           NR_T0 vs R_T0 0.01125320 0.06751921       
#6           NR_T2 vs R_T2 0.23713031 0.35569547 


Shannon_corrected<-phyloseq::estimate_richness(ARGs_ok_corrected, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ok_corrected)) 
Shannon_corrected$Timepoint<-metadata$Timepoint
Shannon_corrected$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon_corrected$Volunteer<-metadata$Volunteer
Shannon_corrected$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon_corrected$GroupTime <-metadata$GroupTime 
# Perform t-tests for the specific pairs
t1 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"])
t2 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])
t3<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "UK"])
t4<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "UK"])

t7 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"])
t8 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t7$p.value, t8$p.value)
)

test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon$signif <- cut(
  test_results_Sannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

test_results_Sannon

#               comparison   p_value     p_adj signif
#1          NR_T0 vs NR_T2 0.3988675 0.7014743       
#2            R_T0 vs R_T2 0.8096220 0.8096220       
#3 NR_T0 Spain vs NR_T0 UK 0.7636711 0.8096220       
#4 NR_T2 Spain vs NR_T2 UK 0.4676495 0.7014743       
#5           NR_T0 vs R_T0 0.1139780 0.5543029       
#6           NR_T2 vs R_T2 0.1847676 0.5543029  
```

###Barplots
```{r barplots ok}
ARGs_ok <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_ok.Rds")

tax<-as.data.frame(tax_table(ARGs_ok))

ARGs_ok<- subset_samples(ARGs_ok, Group_cutoff_4 %in% c("R", "NR"))
ARGs_ok<- subset_samples(ARGs_ok, Timepoint %in% c("T0", "T2"))

library(dplyr)
library(stringr)

tax <- tax %>%
  mutate(Class = case_when(
    str_detect(`Drug Class`, regex("aminoglycoside", ignore_case = TRUE)) ~ "Aminoglycoside",
    str_detect(`Drug Class`, regex("rifamycin", ignore_case = TRUE)) ~ "Rifamycin",
    str_detect(`Drug Class`, regex("tetracycline|minocycline|doxycycline|tigecycline", ignore_case = TRUE)) ~ "Tetracycline",
    str_detect(`Drug Class`, regex("phenicol", ignore_case = TRUE)) ~ "Amphenicol",
    str_detect(`Drug Class`, regex("beta-lactam|penam|penem|monobactam|carbapenem|cephalosporin|cephamycin", ignore_case = TRUE)) ~ "Beta-lactam",
    str_detect(`Drug Class`, regex("diaminopyrimidine|sulfonamide|sulfone|trimethoprim", ignore_case = TRUE)) ~ "Folate pathway antagonist",
    str_detect(`Drug Class`, regex("glycopeptide", ignore_case = TRUE)) ~ "Glycopeptide",
    str_detect(`Drug Class`, regex("ionophore", ignore_case = TRUE)) ~ "Ionophore",  # might not appear in your data
    str_detect(`Drug Class`, regex("lincosamide", ignore_case = TRUE)) ~ "Lincosamide",
    str_detect(`Drug Class`, regex("macrolide", ignore_case = TRUE)) ~ "Macrolide",
    str_detect(`Drug Class`, regex("pleuromutilin", ignore_case = TRUE)) ~ "Pleuromutilin",
    str_detect(`Drug Class`, regex("fluoroquinolone|quinolone", ignore_case = TRUE)) ~ "Quinolone",
    str_detect(`Drug Class`, regex("streptogramin", ignore_case = TRUE)) ~ "Streptogramin",
    str_detect(`Drug Class`, regex("nitroimidazole|nitrofuran", ignore_case = TRUE)) ~ "Nitroimidazole",
    str_detect(`Drug Class`, regex("oxazolidinone|linezolid", ignore_case = TRUE)) ~ "Oxazolidinone",  # rare in this column
    str_detect(`Drug Class`, regex("fosfomycin", ignore_case = TRUE)) ~ "Fosfomycin",
    str_detect(`Drug Class`, regex("elfamycin", ignore_case = TRUE)) ~ "Elfamycin",
    str_detect(`Drug Class`, regex("aminocoumarin", ignore_case = TRUE)) ~ "Aminocoumarin",
    str_detect(`Drug Class`, regex("acridine dye", ignore_case = TRUE)) ~ "Acridine dye",
    str_detect(`Drug Class`, regex("triclosan", ignore_case = TRUE)) ~ "Triclosan",
    str_detect(`Drug Class`, regex("benzalkonium chloride|rhodamine", ignore_case = TRUE)) ~ "Biocide",
    TRUE ~ "Unclassified"
  ))


library(phyloseq)
library(ggplot2)
library(RColorBrewer)

tax$Class[is.na(tax$`Drug Class`)] <- "Unclassified"
tax$Class2 <- tax$Class
tax<-tax[,c("Class", "Class2")]

tax_table(ARGs_ok) <- tax_table(as.matrix(tax))

# Agglomerate at Class level
ARGs_ok_Drug_Class <- tax_glom(ARGs_ok, taxrank = "Class")

# Melt to long format
df_long <- psmelt(ARGs_ok_Drug_Class)


# Step 1: Calculate mean abundance per group and Class
df_summary <- df_long %>%
  group_by(Group_cutoff_4, Class) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

# Step 2: Normalize within each group to make each bar total 100%
df_summary <- df_summary %>%
  group_by(Group_cutoff_4) %>%
  mutate(RelativeAbundance = 100 * Abundance / sum(Abundance)) %>%
  ungroup()

# Step 3: Reorder categories by total relative abundance (optional)
top_categories <- names(sort(tapply(df_summary$RelativeAbundance, df_summary$Class, sum), decreasing = TRUE))
df_summary$Class <- factor(df_summary$Class, levels = top_categories)

# Step 4: Plot

color_palette <- c(
  "Beta-lactam" = "#0072B2",    
  "Folate pathway antagonist" = "#56B4E9",     
  "Lincosamide" = "lightblue",       
  "Tetracycline" = "#DBA5FF",       
  "Rifamycin" = "#8B008B",
  "Quinolone" = "#C8A2C9",    
  "Aminoglycoside" = "#CC79A7",    
  "Fosfomycin"= "#F3D3D9",
  "Amphenicol" = "#D55E00",    
  "Aminocoumarin" = "#E69F00",        
  "Acridine dye" = "#F0E442",    
  "Nitroimidazole" = "#FFFAE5",
  "Macrolide" = "#DBF3C9", 
  "Elfamycin" = "#00C04B",
  "Pleuromutilin" = "#009E73",    
  "Glycopeptide" = "turquoise",    
  "Unclassified" = "#999999"       # grey
)
ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.6) +
  scale_fill_manual(values = color_palette) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)


ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Barplots_ARG_ok.pdf", height = 5, width = 4)


ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)
```

###### Significant diff barplots ARGs
```{r ARGs significance}
library(dplyr)

df_lmm <- df_long %>%
  filter(Timepoint %in% c("T0", "T2"),
         Group_cutoff_4 %in% c("R", "NR")) %>%
  select(Volunteer, Timepoint, Group_cutoff_4, Class, Abundance) %>%
  rename(Phylum = Class)  # so we can reuse the previous run_emmeans_lmm function
emmeans_results <- run_emmeans_lmm(df_lmm)


```




```````` 
not used but saved
#significance taxa
```{r taxa signif}
# to avoid multiple t-testing (lmer + anova)
library(emmeans)
library(lme4)
model<-lmer(Shannon ~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer),data=Shannon)
anova(model)
#Type III Analysis of Variance Table with Satterthwaite's method
#                                 Sum Sq  Mean Sq NumDF  DenDF F value Pr(>F)
#Timepoint                      0.003908 0.003908     1 13.250  0.0290 0.8672
#Group_cutoff_4                 0.017654 0.017654     1 14.017  0.1312 0.7226
#Study                          0.202022 0.202022     1 14.017  1.5016 0.2406
#Timepoint:Group_cutoff_4       0.101950 0.101950     1 13.250  0.7578 0.3995
#Timepoint:Study                0.000773 0.000773     1 13.250  0.0057 0.9407
#Group_cutoff_4:Study           0.003443 0.003443     1 14.017  0.0256 0.8752
#Timepoint:Group_cutoff_4:Study 0.213743 0.213743     1 13.250  1.5887 0.2293



#Fisher
model<-lmer(Fisher ~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer),data=Fisher)
anova(model)
#Type III Analysis of Variance Table with Satterthwaite's method
#                                Sum Sq Mean Sq NumDF  DenDF F value Pr(>F)  
#Timepoint                      262.447 262.447     1 13.067  5.2498 0.0392 *
#Group_cutoff_4                  13.253  13.253     1 13.989  0.2651 0.6147  
#Study                            2.773   2.773     1 13.989  0.0555 0.8172  
#Timepoint:Group_cutoff_4         4.191   4.191     1 13.067  0.0838 0.7767  
#Timepoint:Study                  0.161   0.161     1 13.067  0.0032 0.9556  
#Group_cutoff_4:Study             7.401   7.401     1 13.989  0.1480 0.7062  
#Timepoint:Group_cutoff_4:Study   0.264   0.264     1 13.067  0.0053 0.9432  
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

emm <- emmeans(model, ~ Group_cutoff_4 | Timepoint )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Timepoint = T0:
# contrast estimate   SE   df t.ratio p.value
# NR - R      -5.89 13.2 15.1  -0.446  0.6619

#Timepoint = T2:
# contrast estimate   SE   df t.ratio p.value
# NR - R      -7.42 13.2 15.1  -0.563  0.5818

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Timepoint | Group_cutoff_4 )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Group_cutoff_4 = NR:
# contrast estimate   SE df t.ratio p.value
# T0 - T2      6.84 3.38 13   2.022  0.0642

#Group_cutoff_4 = R:
# contrast estimate   SE df t.ratio p.value
# T0 - T2      5.31 4.08 13   1.300  0.2162

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger  

pairs(emm)
# contrast estimate   SE df t.ratio p.value
# T0 - T2      6.07 2.65 13   2.291  0.0393

#Results are averaged over the levels of: Group_cutoff_4, Study 
#Degrees-of-freedom method: kenward-roger 


#Chao1
model<-lmer(Chao1 ~ Timepoint * Group_cutoff_4 * Study + (1 | Volunteer),data=Chao1)
anova(model)
#Type III Analysis of Variance Table with Satterthwaite's method
#                               Sum Sq Mean Sq NumDF  DenDF F value  Pr(>F)  
#Timepoint                      7498.3  7498.3     1 13.088  4.6922 0.04934 *
#Group_cutoff_4                  509.7   509.7     1 13.989  0.3189 0.58118  
#Study                           522.9   522.9     1 13.989  0.3272 0.57639  
#Timepoint:Group_cutoff_4          0.0     0.0     1 13.088  0.0000 0.99628  
#Timepoint:Study                 542.3   542.3     1 13.088  0.3394 0.57009  
#Group_cutoff_4:Study            444.1   444.1     1 13.989  0.2779 0.60632  
#Timepoint:Group_cutoff_4:Study  215.4   215.4     1 13.088  0.1348 0.71939  
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
 
emm <- emmeans(model, ~ Group_cutoff_4 | Timepoint )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Timepoint = T0:
# contrast estimate   SE   df t.ratio p.value
# NR - R      -36.5 66.5 15.4  -0.549  0.5910

#Timepoint = T2:
# contrast estimate   SE   df t.ratio p.value
# NR - R      -36.7 66.5 15.4  -0.552  0.5891

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 
 
emm <- emmeans(model, ~ Timepoint | Group_cutoff_4 )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
#Group_cutoff_4 = NR:
# contrast estimate   SE df t.ratio p.value
# T0 - T2      32.5 19.1 13   1.701  0.1127

#Group_cutoff_4 = R:
# contrast estimate   SE df t.ratio p.value
# T0 - T2      32.4 23.1 13   1.403  0.1839

#Results are averaged over the levels of: Study 
#Degrees-of-freedom method: kenward-roger 

emm <- emmeans(model, ~ Timepoint  )
#NOTE: Results may be misleading due to involvement in interactions
pairs(emm)
# contrast estimate SE df t.ratio p.value
# T0 - T2      32.5 15 13   2.166  0.0495

#Results are averaged over the levels of: Group_cutoff_4, Study 
#Degrees-of-freedom method: kenward-roger 

# Perform t-tests for the specific pairs
t1 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "NR_T2"])
t2 <- t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])
t3<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "UK"])
t4<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "UK"])
t5<-t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "R_T0" & Shannon$Study == "UK"])
t6<-t.test(Shannon$Shannon[Shannon$GroupTime == "R_T2" & Shannon$Study == "Spain"],
       Shannon$Shannon[Shannon$GroupTime == "R_T2" & Shannon$Study == "UK"])
t7 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T0"])
t8 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
    comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "R_T0 Spain vs R_T0 UK", "R_T2 Spain vs R_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
    p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value, t6$p.value, t7$p.value, t8$p.value)
)

# Adjust p-values using BH method
test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon

#> test_results_Shannon
#               comparison   p_value     p_adj
#1          NR_T0 vs NR_T2 0.6479653 0.8071154
#2            R_T0 vs R_T2 0.7062260 0.8071154
#3 NR_T0 Spain vs NR_T0 UK 0.3507639 0.7115112
#4 NR_T2 Spain vs NR_T2 UK 0.6586883 0.8071154
#5   R_T0 Spain vs R_T0 UK 0.3557556 0.7115112
#6   R_T2 Spain vs R_T2 UK 0.3454075 0.7115112
#7           NR_T0 vs R_T0 0.2652264 0.7115112
#8           NR_T2 vs R_T2 0.8477300 0.8477300

# Perform t-tests for the specific pairs
t1 <- t.test(Fisher$Fisher[Fisher$GroupTime == "NR_T0"],
             Fisher$Fisher[Fisher$GroupTime == "NR_T2"])
t2 <- t.test(Fisher$Fisher[Fisher$GroupTime == "R_T0"],
             Fisher$Fisher[Fisher$GroupTime == "R_T2"])
t3<-t.test(Fisher$Fisher[Fisher$GroupTime == "NR_T0" & Fisher$Study == "Spain"],
           Fisher$Fisher[Fisher$GroupTime == "NR_T0" & Fisher$Study == "UK"])
t4<-t.test(Fisher$Fisher[Fisher$GroupTime == "NR_T2" & Fisher$Study == "Spain"],
           Fisher$Fisher[Fisher$GroupTime == "NR_T2" & Fisher$Study == "UK"])
t5<-t.test(Fisher$Fisher[Fisher$GroupTime == "R_T0" & Fisher$Study == "Spain"],
           Fisher$Fisher[Fisher$GroupTime == "R_T0" & Fisher$Study == "UK"])
t6<-t.test(Fisher$Fisher[Fisher$GroupTime == "R_T2" & Fisher$Study == "Spain"],
           Fisher$Fisher[Fisher$GroupTime == "R_T2" & Fisher$Study == "UK"])
t7 <- t.test(Fisher$Fisher[Fisher$GroupTime == "NR_T0"],
             Fisher$Fisher[Fisher$GroupTime == "R_T0"])
t8 <- t.test(Fisher$Fisher[Fisher$GroupTime == "NR_T2"],
             Fisher$Fisher[Fisher$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Fisher <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "R_T0 Spain vs R_T0 UK", "R_T2 Spain vs R_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value, t6$p.value, t7$p.value, t8$p.value)
)

# Adjust p-values using BH method
test_results_Fisher$p_adj <- p.adjust(test_results$p_value, method = "BH")
test_results_Fisher

#> test_results_Fisher
#               comparison   p_value     p_adj
#1          NR_T0 vs NR_T2 0.4200665 0.8071154
#2            R_T0 vs R_T2 0.7300266 0.8071154
#3 NR_T0 Spain vs NR_T0 UK 0.8703186 0.7115112
#4 NR_T2 Spain vs NR_T2 UK 0.9150340 0.8071154
#5   R_T0 Spain vs R_T0 UK 0.7678223 0.7115112
#6   R_T2 Spain vs R_T2 UK 0.7373593 0.7115112
#7           NR_T0 vs R_T0 0.7459450 0.7115112
#8           NR_T2 vs R_T2 0.5768418 0.8477300

# Perform t-tests for the specific pairs
t1 <- t.test(Chao1$Chao1[Chao1$GroupTime == "NR_T0"],
             Chao1$Chao1[Chao1$GroupTime == "NR_T2"])
t2 <- t.test(Chao1$Chao1[Chao1$GroupTime == "R_T0"],
             Chao1$Chao1[Chao1$GroupTime == "R_T2"])
t3<-t.test(Chao1$Chao1[Chao1$GroupTime == "NR_T0" & Chao1$Study == "Spain"],
           Chao1$Chao1[Chao1$GroupTime == "NR_T0" & Chao1$Study == "UK"])
t4<-t.test(Chao1$Chao1[Chao1$GroupTime == "NR_T2" & Chao1$Study == "Spain"],
           Chao1$Chao1[Chao1$GroupTime == "NR_T2" & Chao1$Study == "UK"])
t5<-t.test(Chao1$Chao1[Chao1$GroupTime == "R_T0" & Chao1$Study == "Spain"],
           Chao1$Chao1[Chao1$GroupTime == "R_T0" & Chao1$Study == "UK"])
t6<-t.test(Chao1$Chao1[Chao1$GroupTime == "R_T2" & Chao1$Study == "Spain"],
           Chao1$Chao1[Chao1$GroupTime == "R_T2" & Chao1$Study == "UK"])
t7 <- t.test(Chao1$Chao1[Chao1$GroupTime == "NR_T0"],
             Chao1$Chao1[Chao1$GroupTime == "R_T0"])
t8 <- t.test(Chao1$Chao1[Chao1$GroupTime == "NR_T2"],
             Chao1$Chao1[Chao1$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Chao1 <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "R_T0 Spain vs R_T0 UK", "R_T2 Spain vs R_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value, t6$p.value, t7$p.value, t8$p.value)
)

# Adjust p-values using BH method
test_results_Chao1$p_adj <- p.adjust(test_results$p_value, method = "BH")
test_results_Chao1

#> test_results_Chao1
#               comparison   p_value     p_adj
#1          NR_T0 vs NR_T2 0.4444955 0.8071154
#2            R_T0 vs R_T2 0.6991097 0.8071154
#3 NR_T0 Spain vs NR_T0 UK 0.9941434 0.7115112
#4 NR_T2 Spain vs NR_T2 UK 0.9975190 0.8071154
#5   R_T0 Spain vs R_T0 UK 0.5659401 0.7115112
#6   R_T2 Spain vs R_T2 UK 0.6513523 0.7115112
#7           NR_T0 vs R_T0 0.6535484 0.7115112
#8           NR_T2 vs R_T2 0.5656997 0.8477300

```


#ARGs rgi
```{r libraries and objects ARGs}
library("phyloseq"); packageVersion("phyloseq")
library(readr)
library(data.table)
library(ggpubr)
library(ggrepel)
library("ggplot2"); packageVersion("ggplot2")
ARGs_rgi_corrected <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_rgi_corrected.Rds")
# Because ARSyN applies log transformation we have negative values, to perform shannon we need no negative entries
otu<-abundances(ARGs_rgi_corrected)+2607523
otu_table(ARGs_rgi_corrected) <- otu_table(otu, taxa_are_rows = TRUE)

ARGs_rgi <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_rgi.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
ARGs_rgi<- subset_samples(ARGs_rgi, Group_cutoff_4 %in% c("R", "NR"))
ARGs_rgi<- subset_samples(ARGs_rgi, Timepoint %in% c("T0", "T2"))

```

```{r ARRs rgi Aplha Diversity}

library(ggsignif)
#Shannon
Shannon<-phyloseq::estimate_richness(ARGs_rgi, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_rgi)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon, aes(x = GroupTime, y = Shannon)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha =1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")
pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_ARGs_rgi.pdf", height = 5, width = 4)


#Shannon corrected
Shannon_corrected<-phyloseq::estimate_richness(ARGs_rgi_corrected, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_rgi_corrected)) 
Shannon_corrected$Timepoint<-metadata$Timepoint
Shannon_corrected$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon_corrected$Volunteer<-metadata$Volunteer
Shannon_corrected$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon_corrected$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon_corrected, aes(x = GroupTime, y = Shannon)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha =1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")
pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_ARGs_rgi_corrected.pdf", height = 5, width = 4)


# Extract OTU table - samples in columns or rows (transpose if needed)
otu_counts <- as.data.frame(t(otu_table(ARGs_rgi)))
# Calculate total reads per sample
total_reads <- rowSums(otu_counts)

# Create data frame and add metadata
metadata <- data.frame(sample_data(ARGs_rgi))
total_reads_df <- data.frame(Sample = rownames(metadata), TotalReads = total_reads)
total_reads_df$Timepoint <- metadata$Timepoint
total_reads_df$Group_cutoff_4 <- metadata$Group_cutoff_4
total_reads_df$Volunteer <- metadata$Volunteer
total_reads_df$Study <- metadata$Study

# Create combined group variable if needed
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
total_reads_df$GroupTime <- metadata$GroupTime

# Plot total reads with ggplot2
pTotalReads <- ggplot(total_reads_df, aes(x = GroupTime, y = TotalReads)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha = 1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Total Reads")
pTotalReads
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Total_reads_ARGrgi.pdf", height = 5, width = 4)

```


```{r comaprisons for significance ARG rgi}
Shannon<-phyloseq::estimate_richness(ARGs_rgi, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_rgi)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
# Perform t-tests for the specific pairs
t1 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "NR_T2"])
t2 <- t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])
t3<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "UK"])
t4<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "UK"])

t7 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T0"])
t8 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t7$p.value, t8$p.value)
)

test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon$signif <- cut(
  test_results_Sannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

test_results_Sannon
#               comparison    p_value     p_adj signif
#1          NR_T0 vs NR_T2 0.41046022
#2            R_T0 vs R_T2 0.78698445
#3 NR_T0 Spain vs NR_T0 UK 0.74853515
#4 NR_T2 Spain vs NR_T2 UK 0.47956947
#5           NR_T0 vs R_T0 0.02870324
#6           NR_T2 vs R_T2 0.26213499


Shannon_corrected<-phyloseq::estimate_richness(ARGs_rgi_corrected, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_rgi_corrected)) 
Shannon_corrected$Timepoint<-metadata$Timepoint
Shannon_corrected$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon_corrected$Volunteer<-metadata$Volunteer
Shannon_corrected$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon_corrected$GroupTime <-metadata$GroupTime 
# Perform t-tests for the specific pairs
t1 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"])
t2 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])
t3<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0" & Shannon_corrected$Study == "UK"])
t4<-t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "Spain"],
           Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2" & Shannon_corrected$Study == "UK"])

t7 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T0"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T0"])
t8 <- t.test(Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "NR_T2"],
             Shannon_corrected$Shannon[Shannon_corrected$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Sannon <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t7$p.value, t8$p.value)
)

test_results_Sannon$p_adj <- p.adjust(test_results_Sannon$p_value, method = "BH")
test_results_Sannon$signif <- cut(
  test_results_Sannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

test_results_Sannon

#               comparison   p_value     p_adj signif
#1          NR_T0 vs NR_T2 0.4104602 0.7193542       
#2            R_T0 vs R_T2 0.7869845 0.7869845       
#3 NR_T0 Spain vs NR_T0 UK 0.7485351 0.7869845       
#4 NR_T2 Spain vs NR_T2 UK 0.4795695 0.7193542       
#5           NR_T0 vs R_T0 0.1161572 0.5226498       
#6           NR_T2 vs R_T2 0.1742166 0.5226498  
```

###Barplots
```{r barplots rgi}
ARGs_rgi <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_rgi.Rds")

tax<-as.data.frame(tax_table(ARGs_rgi))

ARGs_rgi<- subset_samples(ARGs_rgi, Group_cutoff_4 %in% c("R", "NR"))
ARGs_rgi<- subset_samples(ARGs_rgi, Timepoint %in% c("T0", "T2"))

library(dplyr)
library(stringr)

tax <- tax %>%
  mutate(Class = case_when(
    str_detect(`Drug Class`, regex("aminoglycoside", ignore_case = TRUE)) ~ "Aminoglycoside",
    str_detect(`Drug Class`, regex("rifamycin", ignore_case = TRUE)) ~ "Rifamycin",
        str_detect(`Drug Class`, regex("tetracycline|minocycline|doxycycline|tigecycline", ignore_case = TRUE)) ~ "Tetracycline",
    str_detect(`Drug Class`, regex("phenicol", ignore_case = TRUE)) ~ "Amphenicol",
    str_detect(`Drug Class`, regex("beta-lactam|penam|penem|monobactam|carbapenem|cephalosporin|cephamycin", ignore_case = TRUE)) ~ "Beta-lactam",
    str_detect(`Drug Class`, regex("diaminopyrimidine|sulfonamide|sulfone|trimethoprim", ignore_case = TRUE)) ~ "Folate pathway antagonist",
    str_detect(`Drug Class`, regex("glycopeptide", ignore_case = TRUE)) ~ "Glycopeptide",
    str_detect(`Drug Class`, regex("ionophore", ignore_case = TRUE)) ~ "Ionophore",  # might not appear in your data
    str_detect(`Drug Class`, regex("lincosamide", ignore_case = TRUE)) ~ "Lincosamide",
    str_detect(`Drug Class`, regex("macrolide", ignore_case = TRUE)) ~ "Macrolide",
    str_detect(`Drug Class`, regex("pleuromutilin", ignore_case = TRUE)) ~ "Pleuromutilin",
    str_detect(`Drug Class`, regex("fluoroquinolone|quinolone", ignore_case = TRUE)) ~ "Quinolone",
    str_detect(`Drug Class`, regex("streptogramin", ignore_case = TRUE)) ~ "Streptogramin",
    str_detect(`Drug Class`, regex("nitroimidazole|nitrofuran", ignore_case = TRUE)) ~ "Nitroimidazole",
    str_detect(`Drug Class`, regex("oxazolidinone|linezolid", ignore_case = TRUE)) ~ "Oxazolidinone",  # rare in this column
    str_detect(`Drug Class`, regex("fosfomycin", ignore_case = TRUE)) ~ "Fosfomycin",
    str_detect(`Drug Class`, regex("elfamycin", ignore_case = TRUE)) ~ "Elfamycin",
    str_detect(`Drug Class`, regex("aminocoumarin", ignore_case = TRUE)) ~ "Aminocoumarin",
    str_detect(`Drug Class`, regex("acridine dye", ignore_case = TRUE)) ~ "Acridine dye",
    str_detect(`Drug Class`, regex("triclosan", ignore_case = TRUE)) ~ "Triclosan",
    str_detect(`Drug Class`, regex("benzalkonium chloride|rhodamine", ignore_case = TRUE)) ~ "Biocide",
    TRUE ~ "Unclassified"
  ))


library(phyloseq)
library(ggplot2)
library(RColorBrewer)

tax$Class[is.na(tax$`Drug Class`)] <- "Unclassified"
tax$Class2 <- tax$Class
tax<-tax[,c("Class", "Class2")]

tax_table(ARGs_rgi) <- tax_table(as.matrix(tax))

# Agglomerate at Class level
ARGs_rgi_Drug_Class <- tax_glom(ARGs_rgi, taxrank = "Class")

# Melt to long format
df_long <- psmelt(ARGs_rgi_Drug_Class)


# Step 1: Calculate mean abundance per group and Class
df_summary <- df_long %>%
  group_by(Group_cutoff_4, Class) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

# Step 2: Normalize within each group to make each bar total 100%
df_summary <- df_summary %>%
  group_by(Group_cutoff_4) %>%
  mutate(RelativeAbundance = 100 * Abundance / sum(Abundance)) %>%
  ungroup()

# Step 3: Reorder categories by total relative abundance (optional)
top_categories <- names(sort(tapply(df_summary$RelativeAbundance, df_summary$Class, sum), decreasing = TRUE))
df_summary$Class <- factor(df_summary$Class, levels = top_categories)

# Step 4: Plot

color_palette <- c(
  "Beta-lactam" = "#0072B2",    
  "Folate pathway antagonist" = "#56B4E9",     
  "Lincosamide" = "lightblue",       
  "Tetracycline" = "#DBA5FF",       
  "Rifamycin" = "#8B008B",
  "Quinolone" = "#C8A2C9",    
  "Aminoglycoside" = "#CC79A7",    
  "Fosfomycin"= "#F3D3D9",
  "Amphenicol" = "#D55E00",    
  "Aminocoumarin" = "#E69F00",        
  "Acridine dye" = "#F0E442",    
  "Nitroimidazole" = "#FFFAE5",
  "Macrolide" = "#DBF3C9", 
  "Elfamycin" = "#00C04B",
  "Pleuromutilin" = "#009E73",    
  "Glycopeptide" = "turquoise",    
  "Unclassified" = "#999999"       # grey
)
ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.6) +
  scale_fill_manual(values = color_palette) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)


ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Barplots_ARGrgi.pdf", height = 5, width = 4)


ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)
```

#ARGs ResFinder
```{r libraries and objects ARGs}
library("phyloseq"); packageVersion("phyloseq")
library(readr)
library(data.table)
library(ggpubr)
library(ggrepel)
library("ggplot2"); packageVersion("ggplot2")
ARGs_ResFinder <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_ResFinder.Rds")
# Remove Volunteer with less than 1 read from phyloseq object
ARGs_ResFinder<- subset_samples(ARGs_ResFinder, Group_cutoff_4 %in% c("R", "NR"))
ARGs_ResFinder<- subset_samples(ARGs_ResFinder, Timepoint %in% c("T0", "T2"))

```

```{r ARGs ResFinder Aplha Diversity}

library(ggsignif)
#Shannon
Shannon<-phyloseq::estimate_richness(ARGs_ResFinder, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ResFinder)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
pShannon<- ggplot(Shannon, aes(x = GroupTime, y = Shannon)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha =1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Shannon Index")
pShannon
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Alpha_Div_ARGs_ResFinder.pdf", height = 5, width = 4)

# Extract OTU table - samples in columns or rows (transpose if needed)
otu_counts <- as.data.frame(t(otu_table(ARGs_ResFinder)))
# Calculate total reads per sample
total_reads <- rowSums(otu_counts)

# Create data frame and add metadata
metadata <- data.frame(sample_data(ARGs_ResFinder))
total_reads_df <- data.frame(Sample = rownames(metadata), TotalReads = total_reads)
total_reads_df$Timepoint <- metadata$Timepoint
total_reads_df$Group_cutoff_4 <- metadata$Group_cutoff_4
total_reads_df$Volunteer <- metadata$Volunteer
total_reads_df$Study <- metadata$Study

# Create combined group variable if needed
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
total_reads_df$GroupTime <- metadata$GroupTime

# Plot total reads with ggplot2
pTotalReads <- ggplot(total_reads_df, aes(x = GroupTime, y = TotalReads)) +
  geom_boxplot(aes(fill = Group_cutoff_4), position = position_dodge(width = 0.8), alpha = 0.25) +
  geom_jitter(aes(color = Study), alpha = 1,
              position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), size = 2.5) +
  scale_fill_manual(values = c("NR" = "#CC79A7", "R" = "#009E73"), name = "Group cutoff 4") +
  scale_color_manual(values = c("Spain" = "#E69F00", "UK" = "#0072B2"), name = "Cohort") +
  theme_minimal() +
  labs(x = "Group cutoff 4 and Timepoint", y = "Total Reads")
pTotalReads
ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Total_reads_ARGResFinder.pdf", height = 5, width = 4)

```


```{r comaprisons for significance ARG ResFinder}
Shannon<-phyloseq::estimate_richness(ARGs_ResFinder, measures = "Shannon")
metadata<- data.frame(sample_data(ARGs_ResFinder)) 
Shannon$Timepoint<-metadata$Timepoint
Shannon$Group_cutoff_4<-metadata$Group_cutoff_4
Shannon$Volunteer<-metadata$Volunteer
Shannon$Study<-metadata$Study
metadata$GroupTime <- paste(metadata$Group_cutoff_4, metadata$Timepoint, sep = "_")
Shannon$GroupTime <-metadata$GroupTime 
# Perform t-tests for the specific pairs
t1 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "NR_T2"])
t2 <- t.test(Shannon$Shannon[Shannon$GroupTime == "R_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])
t3<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T0" & Shannon$Study == "UK"])
t4<-t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "Spain"],
           Shannon$Shannon[Shannon$GroupTime == "NR_T2" & Shannon$Study == "UK"])

t7 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T0"],
             Shannon$Shannon[Shannon$GroupTime == "R_T0"])
t8 <- t.test(Shannon$Shannon[Shannon$GroupTime == "NR_T2"],
             Shannon$Shannon[Shannon$GroupTime == "R_T2"])

# Combine results into a data frame for p-value adjustment
test_results_Shannon <- data.frame(
  comparison = c("NR_T0 vs NR_T2", "R_T0 vs R_T2", "NR_T0 Spain vs NR_T0 UK", "NR_T2 Spain vs NR_T2 UK", "NR_T0 vs R_T0", "NR_T2 vs R_T2"),
  p_value = c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t7$p.value, t8$p.value)
)

test_results_Shannon$p_adj <- p.adjust(test_results_Shannon$p_value, method = "BH")
test_results_Shannon$signif <- cut(
  test_results_Shannon$p_adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

test_results_Shannon
#               comparison     p_value      p_adj signif
#1          NR_T0 vs NR_T2 0.800496365 0.80049636       
#2            R_T0 vs R_T2 0.364881862 0.43785823       
#3 NR_T0 Spain vs NR_T0 UK 0.043915096 0.08645747       
#4 NR_T2 Spain vs NR_T2 UK 0.057638316 0.08645747       
#5           NR_T0 vs R_T0 0.004644741 0.02786844      *
#6           NR_T2 vs R_T2 0.010484151 0.03145245      *
```

###Barplots
```{r barplots ResFinder}
ARGs_ResFinder <- readRDS("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/ARGs_ResFinder.Rds")
tax<-as.data.frame(tax_table(ARGs_ResFinder))
ARGs_ResFinder<- subset_samples(ARGs_ResFinder, Group_cutoff_4 %in% c("R", "NR"))
ARGs_ResFinder<- subset_samples(ARGs_ResFinder, Timepoint %in% c("T0", "T2"))


library(dplyr)
library(stringr)

tax <- tax %>%
  mutate(Class = case_when(
    str_detect(Phenotype, regex("amikacin|gentamicin|streptomycin|kanamycin|tobramycin|netilmicin|neomycin|dibekacin|sisomicin|spectinomycin", ignore_case = TRUE)) ~ "Aminoglycoside",
    str_detect(Phenotype, regex("chloramphenicol|florfenicol", ignore_case = TRUE)) ~ "Amphenicol",
    str_detect(Phenotype, regex("amoxicillin|ampicillin|cefotaxime|cefoxitin|ceftazidime|piperacillin|ticarcillin|beta-lactam", ignore_case = TRUE)) ~ "Beta-lactam",
    str_detect(Phenotype, regex("trimethoprim|sulfamethoxazole", ignore_case = TRUE)) ~ "Folate pathway antagonist",
    str_detect(Phenotype, regex("vancomycin", ignore_case = TRUE)) ~ "Glycopeptide",
    str_detect(Phenotype, regex("salinomycin|narasin|maduramicin", ignore_case = TRUE)) ~ "Ionophore",
    str_detect(Phenotype, regex("lincomycin|clindamycin", ignore_case = TRUE)) ~ "Lincosamide",
    str_detect(Phenotype, regex("erythromycin|azithromycin|spiramycin|telithromycin", ignore_case = TRUE)) ~ "Macrolide",
    str_detect(Phenotype, regex("tiamulin", ignore_case = TRUE)) ~ "Pleuromutilin",
    str_detect(Phenotype, regex("ciprofloxacin|nalidixic acid|fluoroquinolone", ignore_case = TRUE)) ~ "Quinolone",
    str_detect(Phenotype, regex("pristinamycin IA|quinupristin", ignore_case = TRUE)) ~ "Streptogramin A",
    str_detect(Phenotype, regex("virginiamycin S|virginiamycin M|pristinamycin IIA", ignore_case = TRUE)) ~ "Streptogramin B",
    str_detect(Phenotype, regex("tetracycline|minocycline|doxycycline|tigecycline", ignore_case = TRUE)) ~ "Tetracycline",
    str_detect(Phenotype, regex("metronidazole", ignore_case = TRUE)) ~ "Nitroimidazole",
    str_detect(Phenotype, regex("linezolid", ignore_case = TRUE)) ~ "Oxazolidinone",
    str_detect(Phenotype, regex("rifampin|rifampicin|rifabutin|rifaximin", ignore_case = TRUE)) ~ "Rifamycin",
    TRUE ~ "Unclassified"
  ))

tax$Class[is.na(tax$Phenotype)] <- "Unclassified"
tax$Class2 <- tax$Class
tax$Phenotype<-NULL
tax$`Resistance gene`<-NULL


tax_table(ARGs_ResFinder) <- tax_table(as.matrix(tax))

library(phyloseq)
library(ggplot2)
library(RColorBrewer)

# Agglomerate at Phenotype level
ARGs_ResFinder_Drug_Class <- tax_glom(ARGs_ResFinder, taxrank = "Class")

# Melt to long format
df_long <- psmelt(ARGs_ResFinder_Drug_Class)


# Step 1: Calculate mean abundance per group and Class
df_summary <- df_long %>%
  group_by(Group_cutoff_4, Class) %>%
  summarise(Abundance = mean(Abundance), .groups = "drop")

# Step 2: Normalize within each group to make each bar total 100%
df_summary <- df_summary %>%
  group_by(Group_cutoff_4) %>%
  mutate(RelativeAbundance = 100 * Abundance / sum(Abundance)) %>%
  ungroup()

# Step 3: Reorder categories by total relative abundance (optional)
top_categories <- names(sort(tapply(df_summary$RelativeAbundance, df_summary$Class, sum), decreasing = TRUE))
df_summary$Class <- factor(df_summary$Class, levels = top_categories)

# Step 4: Plot

color_palette <- c(
  "Beta-lactam" = "#0072B2",    # dark blue
  "Tetracycline" = "#56B4E9",     # light blue 
  "Lincosamide" = "lightblue",     # light blue   
  "Aminoglycoside" = "#DBA5FF",    # yellow 
  "Amphenicol" = "#CC79A7",    # pink
  "Macrolide" = "#D55E00",    # dark orange 
  "Folate pathway antagonist" = "#E69F00",        # orange 
  "Nitroimidazole" = "#F0E442",    # green
  "Quinolone" = "lightgreen",    # light green
  "Ionophore" = "#009E73",    # light green
  "Glycopeptide" = "turquoise",    # light light blue
  "Unclassified" = "#999999"       # grey
)
ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.6) +
  scale_fill_manual(values = color_palette) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)

ggsave("/fast/AG_Forslund/Lola/Secuencias_INCLIVA_2024/MHE_rif/outputs/merged/Figures_merged/AlphaDiv/Barplots_ARGresf.pdf", height = 5, width = 4)

ggplot(df_summary, aes(x = Group_cutoff_4, y = RelativeAbundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Relative Abundance (%)", fill = "Class") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 100)

```